Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-11-26 _tex/manuscript.tex
Built site for gh-pages

This assumption makes intuitive sense: The state of the ipsilateral and
contralateral sides of the lymphatic network evolves independently over
time as there are no major lymph vessels crossing the midline. However,
both sides are coupled via time. Qualitatively speaking, according to
the model in \{\#eq-bilateral-marginal\}, a joint state with severe
contralateral involvement and limited ipsilateral involvement would be
unlikely, because severe contralateral involvement becomes likely only
at later time steps when limited ipsilateral involvement becomes
unlikely.

@@

which is less than double in size, compared to the unilateral model.
From these parameters, we now build three different transition matrices:
The unchanged matrix \(\mathbf{A}_\text{i}\) for the ipsilateral side,
one \(\mathbf{A}_\text{c}^{\epsilon=\texttt{False}}\) for the
contralateral side that covers the progression as long as the tumor is
lateralized, and \(\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}\) in
case of a tumor that has crossed the mid-sagittal line.

@@

It can be assumed that most tumors that cross the midline at the time of
diagnose have started as lateralized tumors and grew over the midline at
a later time step. Hence, the transition matrix
\(\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}\) only applies for a
subset of the time steps. Therefore, we also model the tumor's extension
over the mid-sagittal line as a binary random variable. A tumor starts
lateralized and at every time-step there is a finite probability
\(p_\epsilon\) that the tumor grows over the midsaggital plane. The
overall probabilities to find a patient with a clearly lateralized tumor
or one that extends over the mid-sagittal line after \(t\) time-steps
are then respectively given by

@@

The case when midline extension is eventually present is more
complicated: We already marginalize over the exact time-step when the
tumor grows over the mid-sagittal line. But whenever that happens, we
also need to change the contralateral transition matrix to use the
increased spread rates \(b_v^{\text{c}, \epsilon=\texttt{True}}\) from
the tumor to the contralateral LNLs, given by the linear mixing in
equation~\ref{eq-mixing}. We can achieve the correct marginalization by
iteratively building the joint distribution
\(P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t \right)\).
We start from \(t=0\), where we know that all LNLs are healthy (the
contralateral neck is in the state
\(\mathbf{X}_\text{c}=\boldsymbol{\xi}_0\)) and the tumor is lateralized
(\(\epsilon=\texttt{False}\)):

@@

whereas all other states have zero probability. Then we consider an
arbitrary later time-step \(t=\tau+1\). There are two possible scenarios
we need to marginalize over:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The tumor was still lateralized at \(t=\tau\) and just grew over the
  midline. Under this scenario, the probability for having a midline
  extension at \(t=\tau+1\) is given by \(p_\epsilon\). We use this to
  weight the contralateral state distribution that was so far evolved
  without increased contralateral spread.
\item
  The mid-sagittal line was already crossed by the tumor. In that case,
  the probability is 1 to remain in that lateralization state. Thus, to
  account for this case we simply add the distribution
  \(P\left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau \right)\)
  from one time-step earlier, arriving at a recursion relation:
\end{enumerate}

@@

We can use this term to compute the likelihood of all patients with and
without midline extension separately. And if for some patients the
information of tumor lateralization is not available, we can simply sum
the above term once for \(\epsilon=\texttt{False}\) and once for
\(\epsilon=\texttt{True}\) to marginalize over the unknown variable.

@@

In this section, we detail how the experiments were performed. Every
figure, table, and result is fully reproducible via the GitHub
repository
\href{https://github.com/rmnldwg/bilateral-paper}{\texttt{rmnldwg/bilateral-paper}}.
It also contains the raw manuscript and instructions on how to recreate
all figures, tables, and the final document.

@@

It is possible to provide our model with multiple different diagnostic
modalities, each being characterized by different pairs of sensitivity
and specificity. However, we instead chose to combine them into a single
``consensus'' diagnosis before parameter inference, as introduced in
section~\ref{sec-data-consensus}. We opted for this because the
literature values of sensitivity and specificity
\citep{de_bondt_detection_2007, kyzas_18f-fluorodeoxyglucose_2008} of
imaging modalities like MRI and CT do not plausibly match some of our
observations: In the USZ cohort, 78\% of OPSCC patients where diagnosed
with ipsilateral LNL II involvement via diagnostic imaging. This is
virtually impossible with sensitivities around 80\% and specificities
lower than 100\%.

@@

For parameter inference, we used the Python package
\href{https://emcee.readthedocs.io/en/stable/}{\texttt{emcee}}
\citep{foreman-mackey_emcee_2013}. It implements efficient MCMC sampling
algorithms that employ multiple parallel samplers for affine invariance
and better performance on multi-core CPUs. The sample proposal
algirothms used by us are based on differential evolution moves
\citep{ter_braak_differential_2008, nelson_run_2013}. The
\href{https://emcee.readthedocs.io/en/stable/}{\texttt{emcee}} library
was provided with the likelihood implemented by our
\href{https://lymph-model.readthedocs.io/en/stable/}{\texttt{lymph-model}}
Python package.

For each dimension in the parameter space of the model, we initialized
12 of these parallel samplers, called ``walkers'', with random values in
the unit cube. Every time all of these walkers advanced 50 steps, the
autocorrelation time of the chains was estimated. For short chains, this
estimate is not trustworthy, but stabilizes for longer chains. We
therefore considered a sampling to be converged when two criteria were
met:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The change in the autocorrelation time was less then 5.0e-2.
\item
  The estimate of the autocorrelation dropped below \(n\) / 50 where
  \(n\) is the length of the chain up to that point.
\end{enumerate}

All samples up to this convergence - called the \emph{burn-in phase} -
were discarded. We only kept another 10 samples after that, which were
spaced 10 steps apart.

First, in figure~\ref{fig-model-burnin-history}, we verify the sampling
converged successfully by inspecting two monitoring quantities: The
autocorrelation time of the MCMC chain and the acceptance fractions of
the parallel walkers.

@@

\caption{\label{fig-model-burnin-history}Monitoring quantities during
the burn-in phase of the parameter sampling. Left: The autocorrelation
time of the sampling chain estimated at different sampling steps. We
consider the chain converged when the estimate of the autocorrelation
time is stable and drops below the trust threshold of \(n/50\) where
\(n\) is the number of steps. Right: Fraction of accepted MCMC proposals
averaged over all parallel walkers. Values around 30\% indicate good
mixing of the walkers.}

@@

We want to assess the model's capability to approximate the distribution
of lymphatic involvement patterns seen in the data. To that end, we
compare the prevalence of some invovlement patterns under selected
scenarios with the model's prediction for how often these involvements
it expects to see, given these scenarios.

In this context, a ``scenario'' includes the patient's T-category
\(\text{T}x\) and whether the patient's tumor extended over the
mid-sagittal line, i.e.~\(\epsilon=\texttt{True}\) or
\(\epsilon=\texttt{False}\).

An involvement pattern specifies for each ipsi- and contralateral LNL
whether it is ``healthy'', ``involved'', or ``masked''. If it is
``masked'', we essentially state that we are not interested in the
involvement of that LNL and the prevalence will be marginalized over
this LNL's involvement.

For example, we may be interested in the prevalence of contralateral LNL
II involvement (i.e., contra LNL II ``involved'' and all other LNLs
``masked'') under the scenario of early T-category (T0-T2) and no
midline extension (\(\epsilon=\texttt{False}\)). To compute this
prevalence in the data, we select all patients of this scenario (in our
data, this amounts to 379 patients). Of those, 27 were found to harbor
metastases in their contralateral LNL II. Therefore, the prevalence is
7.12\%.

When displaying this data prevalence, we often choose to draw a
\emph{beta posterior} over the ``true'' prevalence, hinting at the fact
that our data merely represents a limited sample. The beta posterior
follows from a uniform beta distribution as prior and a binomial
likelihood for the number of patients with the involvement of interest,
given the parameer for the ``true'' prevalence. The resulting
distribution has its maximum at the observed prevalence, but in addition
gives a visual intuition for the variance of of the observed quantity.
I.e., when we observe 3 out of 10 events, the beta posterior is much
wider than if we observe 300 out of 1000 for he same prevalence. It also
allows us to check not only if the model is accurate, but also whether
it reflects the uncertainty contained in the data.

Predicting the prevalence using our model amounts to computing the
following probability:

@@

Since we compare it to the data, which does not report true but only
observed involvement -- although pathologically investigated LNLs may be
as close as possible to the ground truth -- we do not consider
posteriors of the form \(P \left( \mathbf{X} \mid \mathbf{Z} \right)\)
here. Instead, we compute probabilities of observed involvement
\(P \left( \mathbf{Z} \right)\), as in the likelihood
equation~\ref{eq-single-patient-llh}.

When plotted, we usually display histograms over the model's
predictions. Each of their values was computed from a different
parameter set drawn during MCMC sampling, effectively giving us a
distribution over the prevalences. Ideally, the histograms approximate
the location and width of the Beta posteriors when attempting to
describe the data they were trained on.

Note that we decided to omit the y-axis ticks and labels in these
figures over prevalences and risks. The y-axis in these plots measures
the probability density and its numerical values are not intuitively
interpretable. Instead, we occasionally use the freed space to label
e.g.~rows of subplots.

@@

In table~\ref{tbl-midline-params}, we tabulate the mean and standard
deviation of the sampled parameters for the full midline model.
Considering the parameters of the bilateral model that describe the
ipsilateral spread (\(b^i_v\), \(t_{vr}\) and \(p_\text{adv.}\)), we
note that the bilateral model mostly reproduces the parameter values
reported in the earlier publication \citep{ludwig_modelling_2023} for
the unilateral model. The small differences may be mostly due to the
differences in the training dataset. Therefore, the analysis of the
model's capability to adequately describe ipsilateral lymph node
involvement is not repeated here and we focus instead on the analysis of
contralateral involvement. We can further note that the contralateral
spread parameters \(b^c_v\) are small compared to the ipsilateral
parameters \(b^i_v\), reflecting the low prevalence of contralateral
lymph node involvement for lateralized tumors.

@@

\caption{}\label{T_7c6a5}\tabularnewline
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Mid. ext. probability & 8.16\% & ± 0.48\% \\
ipsi: T ➜ I & 2.80\% & ± 0.26\% \\
ipsi: T ➜ II & 34.89\% & ± 1.40\% \\
ipsi: T ➜ III & 5.45\% & ± 0.66\% \\
ipsi: T ➜ IV & 0.94\% & ± 0.18\% \\
ipsi: T ➜ V & 1.83\% & ± 0.22\% \\
ipsi: T ➜ VII & 2.32\% & ± 0.26\% \\
contra: T ➜ I & 0.29\% & ± 0.09\% \\
contra: T ➜ II & 2.46\% & ± 0.29\% \\
contra: T ➜ III & 0.14\% & ± 0.07\% \\
contra: T ➜ IV & 0.19\% & ± 0.08\% \\
contra: T ➜ V & 0.05\% & ± 0.04\% \\
contra: T ➜ VII & 0.50\% & ± 0.17\% \\
Mixing ⍺ & 33.87\% & ± 4.32\% \\
I ➜ II & 62.50\% & ± 16.69\% \\
II ➜ III & 14.23\% & ± 1.64\% \\
III ➜ IV & 15.86\% & ± 1.93\% \\
IV ➜ V & 14.58\% & ± 3.76\% \\
late T-cat. binom. prob. & 44.98\% & ± 1.99\% \\

@@

In this subsection we illustrate parts of the mathematical framework
defined in the previous section. In figure~\ref{fig-model-midext-evo}
(top panel), we visualize the prior distribution over diagnose times
\(P(t)\). By parameter choice, early T-category tumors are diagnosed on
average after 3 time steps. Advanced T-category tumors are diagnosed on
average after 4.5 time steps, because the parameter \(p_\text{adv.}\)
was learned to be 44.98\% for advanced T-category tumors. The
probability per time step for the tumor to grow over the midline
(\(p_\epsilon\)) was on average 8.16\%. Using this, one can compute the
probability of midline extension \(P(\epsilon \mid t)\) at a given time
step \(t\) (red line in figure~\ref{fig-model-midext-evo}). The bottom
panel in this figure shows the joint probability \(P(\epsilon, t)\) to
be diagnosed at time step \(t\) with a given state of midline extension
and T-category.

@@

\caption{\label{fig-model-midext-evo}The top panel shows the prior
probability to get diagnosed at time-step \(t\) for early and late
T-category tumors as bars. Also in the top panel, we plot the
conditional probability of the tumor's midline extension
(\(\epsilon=\texttt{True}\)), given the time-step \(t\) as a line plot.
In the bottom panel, we show the joint probability of getting diagnosed
in time-step \(t\) \emph{and} having a tumor that crosses the midline.}

@@

The model introduced in the previous section defines a model of the
joint probability distribution over midline extension and ipsi- and
contralateral lymph node involvement,
\(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)\).
How this quantity is computed is illustrated in
figure~\ref{fig-model-algebra-equation}, visually representing
equation~\ref{eq-bilateral-marginal-algebra}. To keep it manageable to
interpret, this example only considers the LNLs II, III, and IV ipsi-
and contralaterally, resulting in \(2^3 = 16\) distinct states per side
and thus \(2 \times 8 \times 8 = 128\) total states. LNLs I, V, and VII
and the associated spread parameters have been removed from the model
while the remaining parameters are set to the mean values in
table~\ref{tbl-midline-params}.

@@

\caption{\label{fig-model-algebra-equation}Visual representation of
equation~\ref{eq-bilateral-marginal-algebra} for the case of midline
extention. The left and right matrices represent the evolution of the
possible hidden states for the ipsi- and contralateral neck
respectively. In particular, the right matrices describe the evolution
of the contralateral hidden states once for the case of no midline
extension (top) and once for the case where the tumor does extend over
the mid-sagittal line (bottom). In the center, the time-prior for late
T-category is plotted as a diagonal matrix. The result of this matrix
computation is the joint distribution
\(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)\),
visualized in figure~\ref{fig-model-state-dist}}

@@

The matrix on the left side of figure~\ref{fig-model-algebra-equation}
shows the time evolution of the probability distribution over the
ipsilateral involvement state, starting from the healthy state
\([0,0,0]\) with probability one. The two matrices on the right side
show the time evolution of the probability distribution over the
contralateral involvement state, where we need to distinguish the two
cases with and without midline extension. At time step \(t=0\), the
contralateral neck starts from the healthy state \([0,0,0]\) without
midline extension. The matrix in the center shows the time prior for
late T-category. The result of the matrix multiplication is the joint
probability distribution over midline extension and ipsi- and
contralateral lymph node involvement,
\(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right)\)
shown in figure figure~\ref{fig-model-state-dist}. The joint
distribution is shown as two separate heatmaps for the two midline
extension states. The most likely state is a lateralized tumor with
involvement of ipsilateral level II and no contralateral involvement,
which has a probability of approximately 25\%. The second most probable
state is a lateralized tumor with ipsilateral involvement of levels II
and III and no contralateral involvement. The most likely state with
contralateral involvement are tumors with midline extension with
involvement of contralateral level II and ipsilateral level II and III.

@@

\caption{\label{fig-model-state-dist}The full distribution
\(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\text{False} \right)\)
over the ipsi- and contralateral states, as well as midline extension as
computed by a model considering the LNLs II, III, and IV for a late
T-category tumor. Since this is a three-dimensional object with one
dimension being the binary random variable representing midline
extension \(\epsilon\), we show this distribution as separate
two-dimensional heatmaps. These two matrices are the result of
equation~\ref{eq-bilateral-marginal-algebra}, which we visualized as
well in figure~\ref{fig-model-algebra-equation}.}

@@

The bilateral model was designed to fulfil the requirements laid out in
section~\ref{sec-requirements}. Now, we investigate to what extent the
model can quantitatively describe the patterns of lymph node involvement
observed in the dataset. To that end, we compare the model's predictions
for contralateral involvement against observations in the data. This is
done given scenarios that differ in T-category and/or midline extension
and/or ipsilateral involvement.

@@

In figure~\ref{fig-model-prevalences-overall}, we plot the prevalence of
contralateral involvement of the LNLs II, III, and IV for the four
scenarios made up of the possible combinations of early and late
T-category, as well as lateralized and midline extending tumors.

@@

Figure~\ref{fig-model-prevalences-overall} shows that the model is
capable of accurately accounting for the most important risk factors,
i.e.~T-category and midline extension. As observed in the data, the
model predicts that the prevalence of contralateral LNL II involvement
increases from 7.1\% for early T-category lateralized tumors to 39.17\%
when the tumor is of advanced T-category and crosses the mid-sagittal
line. Similarly, the prevalence of contralateral LNL III involvement
increases from around 1.55\% for early T-category lateralized tumors to
almost 14.25\% when the tumor is of advanced T-category and crosses the
mid-sagittal line.

@@

\caption{\label{fig-model-prevalences-upstream}The influence of the
upstream LNL II's involvement on the prevalence of contralateral level
III for the four combinations of tumor lateralization (lateralized or
extending over midline) and T-ctageory (early or advanced). Our model
predictions (histograms) are plotted against the observations in the
data (beta posteriors).}

@@

Figure~\ref{fig-model-prevalences-upstream} illustrates how rarely the
contralateral LNL III harbors metastases when its upstream level II is
healthy. Also, the model can capture this correlation rather well, with
the exception of early T-category tumors that extend over the
mid-sagittal line. As in figure~\ref{fig-model-prevalences-overall} and
figure~\ref{fig-model-prevalences-with-ipsi}, these cases are relatively
rare resulting in a broad distribution over the true prevalence.

@@

In figure~\ref{fig-model-prevalences-with-ipsi} we display the model's
ability to capture the correlation between ipsi- and contralateral
involvement. Shown are marginals of the joint distribution where we
consider contralateral LNL II involvement with simulaneous involvement
of different ipsilateral LNLs. The model has no direct connections
between the two sides and thus no adjustable parameter to specifically
quantify correlations between ipsi- and contralateral involvement.
Metastases in the two sides of the neck are only correlated via the time
of diagnosis. Nevertheless, the model adequately predicts that
contralateral level II involvement is rare in combination with a healthy
ipsilateral neck (green histograms). If ipsilateral LNL II is involved,
contralateral involvement is more frequent.

@@

\caption{\label{fig-model-risks}Histograms over the predicted risk of
occult involvement in contralateral LNL II (left), III (middle), and IV
(right), shown for various combinations of T-category, tumor
lateralization, and clinical LNL diagnoses. All LNLs not explicitly
mentioned in the legend, including the LNL for which the risk of occult
disease was computed, where assumed to be clinically negative
(specificity 76\%, sensitivity 81\%).}

>>>>

This assumption is intuitive: since no major lymph vessels cross the
midline, the ipsilateral and contralateral sides of the lymphatic
network evolve independently over time. However, they are indirectly
coupled through time. For example, a joint state with severe
contralateral involvement and limited ipsilateral involvement is
improbable: Severe contralateral involvement typically occurs at later
time steps, when limited ipsilateral involvement is unlikely.

@@

This results in less than a doubling of parameters compared to the
unilateral model. From these parameters, we construct three transition
matrices: the unchanged \(\mathbf{A}_\text{i}\) for the ipsilateral
side, \(\mathbf{A}_\text{c}^{\epsilon=\texttt{False}}\) for
contralateral progression while the tumor is lateralized, and
\(\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}\) for cases where the
tumor crosses the mid-sagittal plane.

@@

Most tumors crossing the midline at the time of diagnosis likely began
as lateralized tumors that grew over the midline at a later stage. As a
result, the transition matrix
\(\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}\) applies only to a
subset of time-steps.

To account for this, we model the tumor's extension over the
mid-sagittal line as a binary random variable. A tumor starts as
lateralized, with a finite probability \(p_\epsilon\) at each time step
of crossing the midline. The overall probabilities of a patient having a
clearly lateralized tumor or one extending over the mid-sagittal line
after \(t\) time steps are then given by

@@

The case of midline extension is more complex: we already marginalize
over the exact time step when the tumor grows over the mid-sagittal
line. However, at the point of crossing, the contralateral transition
matrix must switch to the increased spread rates,
\(b_v^{\text{c}, \epsilon=\texttt{True}}\), as defined by the linear
mixing in equation~\ref{eq-mixing}. To correctly perform this
marginalization, we iteratively construct the joint distribution
\(P \left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid t \right)\).

We begin at \(t=0\), where all contralateral LNLs are healthy (i.e.,
\(\mathbf{X}_\text{c}=\boldsymbol{\xi}_0\)) and the tumor is lateralized
(\(\epsilon=\texttt{False}\)):

@@

while all other states have zero probability.

At some later time step \(t=\tau+1\), there are two scenarios to
marginalize over:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \textbf{The tumor was lateralized at \(t=\tau\) and grew over the
  midline at \(t=\tau+1\):}\\
  In this case, the probability of midline extension at \(t=\tau+1\) is
  \(p_\epsilon\). This probability weights the contralateral state
  distribution that had previously evolved without increased
  contralateral spread.
\item
  \textbf{The tumor had already crossed the midline before
  \(t=\tau\):}\\
  Here, the tumor remains in the midline-crossed state with probability
  1. To account for this scenario, we simply include the distribution
  \(P\left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau \right)\)
  from the previous time step.
\end{enumerate}

Combining these scenarios leads to a recursive formulation:

@@

With the above, we compute the likelihood of all patients with and
without midline extension separately. And if for some patients the
information of tumor lateralization is not available, we can simply
marginalize over the unknown variable
\(\epsilon \in \{ \texttt{False}, \texttt{True} \}\).

@@

This section details the experimental setup. All figures, tables, and
results are fully reproducible via the GitHub repository
\href{https://github.com/rmnldwg/bilateral-paper}{rmnldwg/bilateral-paper}.

@@

Rather than providing the model with a patient's diagnoses from multiple
modalities with different sensitivities and specificities (e.g.~PET-CT
\emph{and} MRI), we combined them into a single ``consensus'' diagnosis
before inference. This approach was necessary because
literature-reported diagnostic imaging sensitivities and specificities
did not plausibly match our observations. For instance, in the USZ
cohort, 78\% of OPSCC patients were diagnosed with ipsilateral LNL II
involvement -\/--- a finding difficult to reconcile with typical imaging
sensitivities around 80\% and specificities below 100\%
\citep{debondt_detection_2007, kyzas_18ffluorodeoxyglucose_2008}.

@@

We used the Python package
\href{https://emcee.readthedocs.io/en/stable/}{\texttt{emcee}}
\citep{foreman-mackey_emcee_2013} for parameter inference, implementing
efficient MCMC sampling with parallel affine-invariant samplers. The
sampling algorithms employed differential evolution moves
\citep{terbraak_differential_2008, nelson_run_2013}, with the likelihood
implemented by our
\href{https://lymph-model.readthedocs.io/en/stable/}{\texttt{lymph-model}}
Python package.

We initialized 12 parallel samplers (``walkers'') with random values
from the unit cube, effectively representing a uniform prior
distribution over the model parameters. Convergence was determined by
two criteria:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The change in autocorrelation time was less than 5.0e-2.
\item
  The autocorrelation estimate dropped below \(n\) / 50, where \(n\) is
  the chain length. Earlier autocorrelation estimates might not be
  trustworthy.
\end{enumerate}

Samples from this \emph{burn-in phase} before convergence were
discarded. After that, we drew 10 additional samples, spaced 10 steps
apart.

We verified sampling convergence in
figure~\ref{fig-model-burnin-history} by examining the MCMC chain's
autocorrelation time and walker acceptance fractions.

@@

\caption{\label{fig-model-burnin-history}Burn-in phase monitoring of
MCMC sampling. Left: Estimated autocorrelation time, indicating
converging when stable and below the trust threshold. Right: Average
acceptance fraction of parallel walkers, with \textasciitilde30\%
indicating good mixing.}

@@

We evaluate the model's ability to approximate the observed distribution
of lymphatic involvement patterns. We compare the prevalence of selected
involvement patterns in the data to the model's predicted prevalence,
given patient scenarios.

A ``scenario'' includes the patient's T-category \(\text{T}x\) and
whether the tumor extended over the mid-sagittal line,
i.e.~\(\epsilon=\texttt{True}\) or \(\epsilon=\texttt{False}\). An
involvement pattern specifies all ipsi- and contralateral LNLs' status
as ``healthy'', ``involved'', or ``masked'' (ignored).

For example, we may assess contralateral LNL II involvement prevalence
under early T-category (T0-T2) and no midline extension
(\(\epsilon=\texttt{False}\)). In the data, 379 such patients were
observed, with 27 exhibiting contralateral LNL II involvement -- a
prevalence of 7.12\%.

To visualize the data prevalence, we plot a \emph{beta posterior} over
the true prevalence \(q\). It is the result of a uniform beta prior over
\(q\), multiplied with the binomial likelihood for \(k\) out of \(n\)
patients, given \(q\). The resulting distribution has its maximum at
\(q=k / n\) and nicely captures the uncertainty in the observed cohort:
if only 3 out of 10 patients are observed with a specific pattern, the
distribution is much wider than if we observe 300 out of 1000.

The model's predicted prevalence to compare it with is computed as:

@@

We display model predictions as histograms, each value computed from one
of the MCMC samples. Ideally, these approximate the location and width
of the beta posteriors from the data showing an accurate and precise
fit.

Note that we omit the y-axis in these figures, as their numerical value
is not intuitively interpretable. We instead use the free space to label
e.g.~rows in an array of subplots.

@@

In table~\ref{tbl-midline-params}, we tabulate the mean and standard
deviation of the sampled parameters. The bilateral model mostly
reproduces the ipsilateral spread parameter values reported in the
earlier publication on the unilateral model
\citep{ludwig_modelling_2023}. Any discrepancies may be due to
differences in the patient cohorts. Therefore, we omit the analysis of
the ipsilateral spread patterns and focus instead on the analysis of
contralateral involvement. Lastly, note the relatively small
contralateral spread parameters \(b^c_v\) compared to the ipsilateral
parameters \(b^i_v\). This accurately reflects the low prevalence of
contralateral lymph node involvement for lateralized tumors.

@@

\caption{}\label{T_d0076}\tabularnewline
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Mid. ext. probability & 8.16\% & ± 0.48\% \\
ipsi: T ➜ I & 2.80\% & ± 0.26\% \\
ipsi: T ➜ II & 34.89\% & ± 1.40\% \\
ipsi: T ➜ III & 5.45\% & ± 0.66\% \\
ipsi: T ➜ IV & 0.94\% & ± 0.18\% \\
ipsi: T ➜ V & 1.83\% & ± 0.22\% \\
ipsi: T ➜ VII & 2.32\% & ± 0.26\% \\
contra: T ➜ I & 0.29\% & ± 0.09\% \\
contra: T ➜ II & 2.46\% & ± 0.29\% \\
contra: T ➜ III & 0.14\% & ± 0.07\% \\
contra: T ➜ IV & 0.19\% & ± 0.08\% \\
contra: T ➜ V & 0.05\% & ± 0.04\% \\
contra: T ➜ VII & 0.50\% & ± 0.17\% \\
Mixing ⍺ & 33.87\% & ± 4.32\% \\
I ➜ II & 62.50\% & ± 16.69\% \\
II ➜ III & 14.23\% & ± 1.64\% \\
III ➜ IV & 15.86\% & ± 1.93\% \\
IV ➜ V & 14.58\% & ± 3.76\% \\
late T-cat. binom. prob. & 44.98\% & ± 1.99\% \\

@@

In this subsection, we illustrate key aspects of the mathematical
framework introduced earlier. The top panel of
figure~\ref{fig-model-midext-evo} shows the prior distribution over
diagnosis times, \(P(t)\). Based on the parameterization, early
T-category tumors are on average diagnosed after 3 time steps, while
advanced T-category tumors are diagnosed later, averaging 4.5 time
steps. This is due to the learned value of \(p_\text{adv.}\), which is
44.98\% for advanced T-category tumors. The tumor's average probability
per time step of growing over the midline, \(p_\epsilon\), was found to
be 8.16\%. Using this value, the conditional probability of midline
extension, \(P(\epsilon \mid t)\), can be computed for a given time step
\(t\) (red line in the top panel of figure~\ref{fig-model-midext-evo}).
The bottom panel visualizes the joint probability \(P(\epsilon, t)\),
showing the likelihood of diagnosis at time \(t\) with specific states
of midline extension and T-category.

@@

\caption{\label{fig-model-midext-evo}The top panel shows the prior
probability to be diagnosed at time step \(t\) for early and late
T-category tumors as bars. The conditional probability of midline
extension (\(\epsilon=\texttt{True}\)) given time step \(t\) is shown as
a line plot. The bottom panel illustrates the joint probability of being
diagnosed at time \(t\) and having a tumor that crosses the midline.}

@@

The framework models the joint probability distribution of midline
extension and ipsi- and contralateral lymph node involvement,
\(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)\).
This is visualized in figure~\ref{fig-model-algebra-equation}, which
represents the calculation defined in
equation~\ref{eq-bilateral-marginal-algebra}. To simplify
interpretation, the example focuses only on LNLs II, III, and IV,
reducing the state space to \(2^3 = 16\) possible states per side, and
\(2 \times 8 \times 8 = 128\) total states. LNLs I, V, and VII are
excluded, along with their spread parameters, while remaining parameters
are set to their mean values from table~\ref{tbl-midline-params}.

@@

\caption{\label{fig-model-algebra-equation}Visual representation of
equation~\ref{eq-bilateral-marginal-algebra} for midline extension. The
left and right matrices represent the time evolution of hidden states
for the ipsi- and contralateral necks, respectively. The right matrices
distinguish between the cases of no midline extension (top) and midline
extension (bottom). The central diagonal matrix shows the time-prior for
late T-category tumors. This computation yields the joint distribution
\(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)\),
visualized in figure~\ref{fig-model-state-dist}.}

@@

The left matrix in figure~\ref{fig-model-algebra-equation} shows the
time evolution of the probability distribution over the ipsilateral
involvement states, starting from the healthy state \([0,0,0]\). The two
right matrices show the contralateral state evolution, distinguishing
between the cases of midline extension and no midline extension. At
\(t=0\), the contralateral neck begins in the healthy state \([0,0,0]\)
without midline extension. The central matrix shows the time prior for
late T-category tumors. The matrix multiplication results in the joint
distribution
\(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right)\),
visualized in figure~\ref{fig-model-state-dist}.

This joint distribution is presented as two heatmaps, corresponding to
the two states of midline extension. The most likely state involves a
lateralized tumor with ipsilateral level II involvement and no
contralateral involvement, having a probability of approximately 25\%.
The next most probable state involves ipsilateral levels II and III
without contralateral involvement. The most likely state with
contralateral involvement corresponds to tumors with midline extension,
showing involvement of contralateral level II and ipsilateral levels II
and III.

@@

\caption{\label{fig-model-state-dist}The full distribution
\(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\text{False} \right)\)
over ipsi- and contralateral states and midline extension for late
T-category tumors. This three-dimensional distribution is shown as two
separate heatmaps for the binary midline extension variable
\(\epsilon\). These matrices are the result of
equation~\ref{eq-bilateral-marginal-algebra}, visualized in
figure~\ref{fig-model-algebra-equation}.}

@@

The bilateral model was designed to meet the requirements outlined in
section~\ref{sec-requirements}. Here, we evaluate the model's ability to
quantitatively capture the observed patterns of lymph node involvement
in the dataset. Specifically, we compare the model's predictions for
contralateral involvement to the observed data across scenarios that
vary by T-category, midline extension, and ipsilateral involvement.

@@

In figure~\ref{fig-model-prevalences-overall}, we compare the prevalence
of contralateral involvement for LNLs II, III, and IV across four
scenarios: combinations of early and late T-category, and lateralized
versus midline-extending tumors.

@@

Figure~\ref{fig-model-prevalences-overall} demonstrates the model's
ability to account for key risk factors, such as T-category and midline
extension. Consistent with the data, the model predicts that the
prevalence of contralateral LNL II involvement increases from 7.1\% for
early T-category lateralized tumors to 39.17\% for advanced T-category
tumors that cross the midline. Similarly, contralateral LNL III
involvement rises from around 1.55\% for early T-category lateralized
tumors to nearly 14.25\% for advanced T-category midline-extending
tumors.

@@

\caption{\label{fig-model-prevalences-upstream}The influence of the
upstream LNL II's involvement on the prevalence of contralateral level
III for the four combinations of tumor lateralization (lateralized or
extending over midline) and T-category (early or advanced). Our model
predictions (histograms) are plotted against the observations in the
data (beta posteriors).}

@@

Figure~\ref{fig-model-prevalences-upstream} highlights the influence of
upstream LNL II involvement on contralateral LNL III metastasis. The
contralateral LNL III rarely harbors metastases when its upstream LNL II
is healthy, a correlation well-captured by the model. However, for early
T-category tumors extending over the midline, the model slightly
underestimates the observed prevalence, likely due to the relative
rarity of such cases, resulting in broader uncertainty in the observed
prevalence estimates.

@@

In figure~\ref{fig-model-prevalences-with-ipsi}, the model's ability to
capture the correlation between ipsi- and contralateral involvement is
demonstrated. The marginals of the joint distribution highlight
contralateral LNL II involvement alongside varying ipsilateral LNL
involvement states. Despite having no direct connections between the two
sides, the model successfully predicts these correlations, which arise
purely through the shared diagnosis time.

For example, the model accurately predicts that contralateral LNL II
involvement is rare when the ipsilateral neck is completely healthy
(green histograms). However, if ipsilateral LNL II is involved,
contralateral involvement becomes more likely. Notably, the model
achieves this without specific parameters to quantify ipsi-contralateral
correlations, relying instead on the inherent structure of the
time-dependent dynamics.

@@

\caption{\label{fig-model-risks}Histograms over the predicted risk of
occult involvement in contralateral LNL II (left), III (middle), and IV
(right), shown for various combinations of T-category, tumor
lateralization, and clinical LNL diagnoses. All LNLs not explicitly
mentioned in the legend, including the LNL for which the risk of occult
disease was computed, were assumed to be clinically negative
(specificity 76\%, sensitivity 81\%).}
